The low-frequency environment of the Murchison Widefield Array: radio-frequency interference analysis and mitigation

### 2014-11-17 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Implemented Emil's comments

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be built near the location of the MWA.

Several levels of protection are in place to protect the radio quietness of the MRO. Within a 70~km radius, the Australian Communications and Media Authority\footnote{\url{http://www.acma.gov.au/}} and WA government provide the strongest level of protection from other radio equipment across the frequency range 70~MHz to 25.25~GHz. Radio devices in this zone must not cause interference to radio astronomy.  Beyond 70~km, coordination zones extend out to~260 km radius at the lowest frequencies and reduce in size with increasing frequency. Individually licensed or spectrum-licensed radio devices in these zones must be coordinated with radio-astronomy requirements to eliminate or minimize interference. Situations not covered by the radio-quiet zone are: i) transmitters below 70~MHz; ii) most aircraft transmissions; iii) satellite transmissions; and iv) transmitters beyond 260~km from the centre of the MRO.

@@

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size and without increasing reading and writing it requires 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time segments and flags those independently. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. As not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been set up that pre-runs \textsc{cotter} on all observations. In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.

@@

Fig.~\ref{fig:rfi-per-band} shows the overall detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17. The latter is the only night affected by interference from digital TV (DTV), and will be analysed later. The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data. Excluding the RFI from DTV, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy. The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.

@@

The \textsc{aoflagger} RFI detection strategy, originally developed for LOFAR, works well for the MWA. Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events. These algorithms have gained some popularity; besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in \textsc{miriad}. Nevertheless, other projects still use single-sample thresholding, e.g. PAPER \citep{parsons-paper-eorlimit-2014}. Since strong RFI is seen practically everywhere, and because the apparent strength of RFI events will follow a power-law distribution, many faint transmitters will interfere with observations of any (terrestrial) radio observatory. Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in long-integration projects.

>>>>

In this article, we will look specifically at the RFI situation of the Murchison Widefield Array (MWA; \citealt{mwa-design-2009,mwa}). The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarised dipoles, which allow observing between 72--300~MHz with a 30.72-MHz instantaneous bandwidth; one of its main science drivers is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located at the CSIRO Murchison Radio-astronomy Observatory (MRO) in the Murchison region of Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the SKA, because the cores of the SKA low-frequency aperture array are planned to be built near the location of the MWA.

Several levels of protection are in place to protect the radio quietness of the MRO. Within a 70~km radius, the Australian Communications and Media Authority\footnote{\url{http://www.acma.gov.au/}} and Western Australia government provide the strongest level of protection from other radio equipment across the frequency range 70~MHz to 25.25~GHz. Radio devices in this zone must not cause interference to radio astronomy.  Beyond 70~km, coordination zones extend out to~260 km radius at the lowest frequencies and reduce in size with increasing frequency. Individually licensed or spectrum-licensed radio devices in these zones must be coordinated with radio-astronomy requirements to eliminate or minimize interference. Situations not covered by the radio-quiet zone are: i) transmitters below 70~MHz; ii) most aircraft transmissions; iii) satellite transmissions; and iv) transmitters beyond 260~km from the centre of the MRO.

@@

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The SumThreshold and SIR operator algorithms that are used by the flagging strategy use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size and without increasing reading and writing it requires 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time segments and flags those independently. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. As not all astronomers have easy access to large-memory machines, a platform with sufficient memory has been set up that runs a partial \textsc{cotter} preprocessing on all observations. In this use-case, \textsc{cotter} is run in a flagging-only mode on the large-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.

@@

Fig.~\ref{fig:rfi-per-band} shows the overall detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17. The latter is the only night affected by interference from digital TV (DTV), and will be analysed later in this section. The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data. Excluding the RFI from DTV, the EoR high band is slightly cleaner than the EoR low band, and its worst subband at 188.2 MHz has 1.03\% occupancy. The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy.

@@

The \textsc{aoflagger} RFI detection strategy, originally developed for LOFAR, works well for the MWA. Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events. These algorithms have gained some popularity; besides the use of \textsc{aoflagger} by individual astronomers, the SumThreshold algorithm is implemented in the pipeline for eMERLIN \citep{serpent-peck-2013} and in \textsc{miriad}'s PGFLAG task. Nevertheless, other projects still use single-sample thresholding, e.g. PAPER \citep{parsons-paper-eorlimit-2014}. Since strong RFI is seen practically everywhere, and because the apparent strength of RFI events will follow a power-law distribution, many faint transmitters will interfere with observations of any (terrestrial) radio observatory. Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in long-integration projects.
